---
title: "restfulSE -- experiments with HDF5 server content wrapped in SummarizedExperiment"
author: "Vincent J. Carey, stvjc at channing.harvard.edu, Shweta Gopaulakrishnan, reshg at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{restfulSE -- experiments with SE interface to remote HDF5}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::pdf_document:
    toc: yes
    number_sections: yes
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

Extensive human and computational effort is expended on
downloading and managing large genomic data at site of analysis.
Interoperable formats that are accessible via generic operations
like those in RESTful APIs may help to improve cost-effectiveness
of genome-scale analyses.

In this report we examine the use of HDF5 server as a back end
for assay data, mediated through the RangedSummarizedExperiment API
for interactive use.

A modest server configured to deliver HDF5 content via a
RESTful API has been prepared and is used in this vignette.

# Background

Banovich et al. published a subset of DNA methylation measures
assembled on 64 samples of immortalized B-cells from the YRI HapMap cohort.

```{r setup,echo=FALSE}
suppressPackageStartupMessages({
library(restfulSE)
library(SummarizedExperiment)
})
```{r lkd}
library(restfulSE)
data(banoSEMeta)
banoSEMeta
```

The numerical data have been exported using H. Pages' 
saveHDF5SummarizedExperiment applied to the banovichSE
SummarizedExperiment in the yriMulti package.  The HDF5
component is simply copied into the server data space
on the remote server.  A wrapper class has been defined
in the restfulSE package.

```{r lkass}
banoh5  = banoH5("http://54.163.220.201:5000")
banoh5
```

# Construction

```{r docon}
restBano = RESTfulSummarizedExperiment(banoSEMeta, banoh5)
restBano
```

# Subsetting and assay extraction

Targeted extraction is possible, but index processing needs
considerable work.

```{r dosub}
subr = restBano[1:4,1:6]
subr
assay(subr)
```
