---
title: "restfulSE -- experiments with HDF5 server content wrapped in SummarizedExperiment"
author: "Vincent J. Carey, stvjc at channing.harvard.edu, Shweta Gopaulakrishnan, reshg at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{restfulSE -- experiments with SE interface to remote HDF5}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::pdf_document:
    toc: yes
    number_sections: yes
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

Extensive human and computational effort is expended on
downloading and managing large genomic data at site of analysis.
Interoperable formats that are accessible via generic operations
like those in RESTful APIs may help to improve cost-effectiveness
of genome-scale analyses.

In this report we examine the use of HDF5 server as a back end
for assay data, mediated through the RangedSummarizedExperiment API
for interactive use.

A modest server configured to deliver HDF5 content via a
RESTful API has been prepared and is used in this vignette.

# Executive summary

We want to provide rapid access to array-like data.  We'll work
with the Banovich 450k data as there is a simple check against
an in-memory representation.

```{r setup,echo=FALSE}
suppressPackageStartupMessages({
library(restfulSE)
library(SummarizedExperiment)
})
```{r dobanoyy}
library(restfulSE)
bigec2 = H5S_source("http://54.174.163.77:5000")
bigec2
dsmeta(bigec2)[1:2,] # two groups
dsmeta(bigec2)[1,2][[1]] # all dataset candidates in group 1
```

We use double-bracket subscripting to grab a reference to a
dataset from an H5S source.
```{r doba2}
banref = bigec2[["assays"]] # arbitrary name assigned long ago
banref
```

We build a RESTfulSummarizedExperiment by combining an assay-free
RangedSummarizedExperiment with this reference.
```{r doba3}
data(banoSEMeta)
rbano = RESTfulSummarizedExperiment(banoSEMeta, banref)
rbano
```

We can update the SummarizedExperiment metadata 
through subsetting operations, and then extract the relevant
assay data.  The data are retrieved from the remote server
with the `assay` method.
```{r doba4}
rbanoSub = rbano[5:8, 3:9] # currently only trivial subsets retrieved
assay(rbanoSub) # general index processing under construction
```
   

# Background

Banovich et al. published a subset of DNA methylation measures
assembled on 64 samples of immortalized B-cells from the YRI HapMap cohort.

```{r lkd}
library(restfulSE)
data(banoSEMeta)
banoSEMeta
```

The numerical data have been exported using H. Pages' 
saveHDF5SummarizedExperiment applied to the banovichSE
SummarizedExperiment in the yriMulti package.  The HDF5
component is simply copied into the server data space
on the remote server.  


# Hierarchy of server resources

## Server

Given the URL of a server running HDF5 server, we create
an instance of `H5S_source`:
```{r doso}
mys = H5S_source(serverURL="http://54.163.220.201:5000")
mys
```

## Groups

The server identifies a collection of 'groups'.  For the server
we are working with, only one group, at the root, is of interest.
```{r lkg}
groups(mys)
```

## Links for a group

There is a class to hold the link set for any group:
```{r lkhl}
lin1 = links(mys,1)
lin1
```

